---
<<<<<<< Updated upstream
title: "Exploratory Data Analysis"
output:
  html_document:
    df_print: paged
=======
title: "Exploratory Data Analysis "
author: "Kaustubh Saraf"
date: "`r Sys.Date()`"
output: html_document
>>>>>>> Stashed changes
---

## EDA Using R {.tabset .tabset-fade .tabset-pills}

### Packages Required {.tabset .tabset-fade .tabset-pills}

1. tidyverse : This package consists of 6 core packages out of which the below 3 are most important for this project: 
 * ggplot2: Used for creating powerful visualizations
 * dplyr: Used for data manipulation
 * tidyr: Used for data modifications
 
2. rpart, rpart.plot and ROCR : These packages are used for building classification and regression models using decision trees. Further, we can visualize the tree structure and evaluate the performance of the models 

3. forecast, tseries and sarima : These packages are used to model the time-series data including the seasonal component in the series if any


```{r warning=FALSE, message=FALSE}
library(tidyverse) #used for data manipulation
library(rmarkdown) #used for formatting the markdown file

library(sqldf) #using SQL commands in R
```



### Data Preparation {.tabset .tabset-fade .tabset-pills}

```{r}
CustomerData<-read.csv("data/CustomerData.csv",header=TRUE,sep=",")
```

**1.1 Take a look into the data**
```{r}
head(CustomerData)
```

**1.2 Size of the dataset**
**1.2.1 How larger is my dataset? How many rows? How many columns?**
```{r}
dim(CustomerData)
```

**1.2.2 How many columns?**
```{r}
ncol(CustomerData)
```

**1.2.3 How many rows?**
```{r}
nrow(CustomerData)
```

**1.3 Glimpse at the variables**

**1.3.1 What is the name of my columns?**
```{r}
names(CustomerData)

#OR

colnames(CustomerData)
```

**1.3.2 What is the structure of the data**
```{r}
str(CustomerData)
```

**1.3.3 How to get the format a column**
```{r}
class(CustomerData[,1])
```

**1.4 Summary statistics of the dataset**
**1.4.1 Getting the summary of the dataset**
```{r}
summary(CustomerData)
```

**1.4.2 Getting some other statistics**
```{r}
#Standard Deviation
sd(CustomerData$Age)

#Quantiles
quantile(CustomerData$Age)
```

**1.4.3 Getting statistics for multiple columns**
```{r}
apply(CustomerData[,c(5,6,11)], 2, sd) 
```

**1.4.4 Summarizing the dataset to aggregate data and calculate various statistical values**
```{r}
#Mean
aggregate(.~Gender, CustomerData, mean) 

#Std. Deviation
aggregate(.~Gender, CustomerData, sd)
```

**1.4.5 One-way counting**
```{r}
table(CustomerData$Gender)
```

**1.4.6 Two way counting table**
```{r}
table(CustomerData$Gender,CustomerData$UnionMember)
```


### EDA using Visualizations {.tabset .tabset-fade .tabset-pills}

**2.1.1 Single Histogram**
```{r}
hist(CustomerData$Age, col="yellow", breaks=20)
```

**2.1.2 Histogram combined with density curve**
```{r}
hist(CustomerData$Age, prob=T, col="yellow", breaks=20, main="Histogram and Density of Age", xlim=c(min(CustomerData$Age)-1,max(CustomerData$Age)+1), xlab="Age")
lines(density(CustomerData$Age), col="red", lwd=2)

# Add a vertical line that indicates the average of Sepal Length
abline(v=mean(CustomerData$Age), col="blue", lty=2, lwd=1.5)
```

* You can change the color of the histogram, the density curve and the vertical line
* In addition to that, you can set the range of your x-axis by varying values in xlim
* Further, you can use median or other aggregations to plot the vertical line


**2.1.3 Plotting multiple bar charts**
```{r}
avg<- apply(CustomerData[,c(5,6)], 2, mean)
barplot(avg, ylab = "Average")
```

**2.1.4 Plotting multiple bar charts for same variable split by another variable**
```{r}
counts <- table(CustomerData$UnionMember,CustomerData$Gender)
barplot(counts, main="Union Members by Gender",
  xlab="Gender", col=c("darkblue","yellow"),
  legend = rownames(counts), beside=TRUE)
```

**2.2 Finding the outliers**

**2.2.1 Single box plot**
```{r}
boxplot(CustomerData$Age)
```

* The center line represents the median and the two ends of the box represent the 25th and 75th percentiles or 1st and 3rd Quantile
* The outer lines represent 1.5 x IQR (Inter Quantile Range is the width of the box)


**2.2.2 Plotting multiple box plots in same graph**
```{r}
boxplot(CustomerData[,c(5,6,9)], notch=T, col=c("blue", "yellow","red"))
```


**2.2.3 Box plot of single variable with groups**
```{r}
boxplot(CustomerData[,'HouseholdSize']~CustomerData[,'Gender'], notch=T, ylab="Household Size", col="blue")
```


**2.3.1 Scatter plot to observe the relation between two variables**
```{r}
plot(CustomerData$Age, CustomerData$HouseholdSize, xlab = "Age", ylab = "HH Size", main = "Household Size vs Age")
```

**2.3.2 Scatter plot of multiple variables **
```{r}
pairs(CustomerData[,c(5,6,11)])
```


**2.4 Plotting multiple graphs in same window**
```{r}
# set arrangement of multiple plots. 2,2 will allow us to plot 4 graphs in a window
par(mfrow=c(2,2))
# set margins
par(mar=c(4.5, 4.2, 3, 1.5)) 

#Graph1
hist(CustomerData$Age, xlab = "Age", cex.lab=1.5, col = "yellow")

#Graph2
hist(CustomerData$HHIncome, xlab = "Household Income", col = "red")

#Graph3 (pch allows us to change shapes)
plot(CustomerData$Age, CustomerData$HouseholdSize, xlab = "Age", ylab = "Household Size", main= "Household Size vs Age", pch=17)

#Graph4
boxplot(CustomerData[,c(5,6)], notch=T, col=c("red", "blue"))
```



### Tidyverse for EDA {.tabset .tabset-fade .tabset-pills}

#### Row Manipulations {.tabset .tabset-fade .tabset-pills}

**3.1 Filtering data**

**3.1.1 Method 1 : Using subset() function**
```{r}
subset(x = CustomerData, subset = Age > 25 & HHIncome > 400000)

# OR

subset(CustomerData, Age > 25 & HHIncome > 400000)

# OR

CustomerData[(CustomerData$Age > 25 & CustomerData$HHIncome > 400000), ]
```


**3.1.2 Method 2 : Using SQL to filter data**
```{r}
sqldf('select * from CustomerData where `Age` > 25 and `HHIncome` > 400000')
```


**3.1.3 Method 3 : Using filter() function**
```{r}
CustomerData1 <- filter(CustomerData, Age>25 & HHIncome>400000)
CustomerData2 <- filter(CustomerData1, Gender=="Female", EducationYears<=20 | UnionMember=="Yes")
```

**3.2.1 Random subsetting of data**
```{r}
# Randomly sampling 90% data
CustomerDataSample1 <- CustomerData[sample(x = nrow(CustomerData), size = nrow(CustomerData)*0.90),]

# OR

CustomerDataSample2 <- sample_frac(CustomerData, 0.9)

# To sample fixed number of rows
CustomerDataSample3 <- sample_n(CustomerData, 1000)

```


**3.2.2 Randomly splitting data into two groups**
```{r}
index <- sample(nrow(CustomerData),nrow(CustomerData)*0.90)
CustomerData.train = CustomerData[index,]
CustomerData.test = CustomerData[-index,]
```


**3.3 Sorting data**
**Here we are extracting the first 5 rows. This can be modified based on the requirements**
```{r}
CustomerData[order(CustomerData$Age, decreasing = TRUE)[1:5], ] 

# OR

arrange(CustomerData, desc(Age))[1:5, ] #Descending

# OR 

arrange(CustomerData, Age)[1:5, ] #Ascending

# Sorting using multiple columns

CustomerData_sort<- arrange(CustomerData, Age, desc(HHIncome))[1:5,]
```


#### Column Manipulations {.tabset .tabset-fade .tabset-pills}

**3.4 Variable details**
**3.4.1 Extracting column names**
```{r}
names(CustomerData)
```

**3.4.2 Selecting columns**
```{r}
CustomerData[, c("Age", "HHIncome")]

# OR

varname <- c("Age", "HHIncome")
CustomerData_select <- select(CustomerData, varname)

# OR

CustomerData_select <- CustomerData[,varname]

```


**3.4.3 Dropping columns**
```{r}
CustomerData_select2 <- select(CustomerData, -Age, -HHIncome)

# OR

varname <- c("Age", "HHIncome")
CustomerData_select2 <- CustomerData[,!names(CustomerData) %in% varname]
names(CustomerData_select2)
```


**3.4.4 Reordering columns**
```{r}
CustomerData_order2 <- select(CustomerData, Age, HHIncome, everything())

names(CustomerData_order2)
```

**3.4.5 Renaming variables**
```{r}
CustomerData_rename<- rename(CustomerData, Education=EducationYears, Income=HHIncome)

CustomerData_rename1<-select(CustomerData_rename, Education, Income, everything())

names(CustomerData_rename1)
```


**3.4.6 Creating new variable**

**There are multiple ways to create new columns. Using mutate is the simplest while other methods might require multiple steps**

```{r}
CustomerData_newvar<- mutate(CustomerData, Age_Income_Ratio=HHIncome/Age)
names(CustomerData_newvar)
```

#### Missing Data Treatment {.tabset .tabset-fade .tabset-pills}

**3.5 Missing data analysis**
**3.5.1 Getting the number of missing values in the dataset**
```{r}
sum(is.na(CustomerData))
```

**3.5.2 Getting the missing values for each variable**
```{r}
colSums(is.na(CustomerData))
```


**3.5.3 Removing entries with missing values**
```{r}
CustomerData.Clean<- na.omit(CustomerData)

sum(is.na(CustomerData.Clean))
```


**3.5.4 Removing entries with missing values for a specific variable**
```{r}
CustomerData.Clean1<-CustomerData[!is.na(CustomerData$NumberPets),]

colSums(is.na(CustomerData.Clean1))
```
